Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-04-26 paper/ms.tex
Add ref to UGC 628
Many cosmetic changes
Moving figures around, adding authors, acknowledgements, etc

\usepackage{epsfig,caption}
\usepackage{color}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{longtable}
% \usepackage{amssymb}
\usepackage{rotating,xcolor}
\usepackage{hyperref}
% \usepackage{tgbonum}
\usepackage[symbol]{footmisc}

@@

\usepackage[misc]{ifsym}
\usepackage{xcolor}

@@

\author{Angus Beane$^{1*}$, Lars Hernquist$^1$, Elena D'Onghia$^{2,3}$, et al.}

@@

\begin{affiliations}
\item Center for Astrophysics $|$ Harvard \& Smithsonian,  Cambridge, MA, USA

\item Department of Physics, University of Wisconsin-Madison, Madison, WI, USA

\item Department of Astronomy, University of Wisconsin-Madison, Madison, WI, USA

@@

@@

The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics.\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R} Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar.\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C} Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence.\cite{2019MNRAS.490.4740B} And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

@@

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which are exceptionally
gas-poor to be rotating very slowly. We therefore predict that observations
which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. In
general, we predict that all gas-rich spiral galaxies should rotate quickly
while gas-poor spiral galaxies should rotate slowly.

\begin{figure*}[h]%
\centering
\includegraphics[width=0.8\textwidth]{fig/fig1.pdf}
\caption{Surface density projections in an N-body only simulation (upper
panels) and a simulation which includes the SMUGGLE model (bottom panels). We
can see that in the N-body run, the bar grows in length and strength. In the
SMUGGLE run (which includes a gaseous phase and a model for the multi-phase
interstellar medium), the bar remains at approximately the same length and
strength over the course of the simulation. The N-body model is identical to
the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

\begin{figure}[h]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig2.pdf}
\caption{Various bar properties from the N-body and SMUGGLE runs. The
\textit{upper panel} show the evolution of the pattern speed. As expected, the
bar in the N-body run slows down due to interactions between the bar and the
dark matter halo. However, the bar in the SMUGGLE run does not slow down and
instead remains at a constant pattern speed. The \textit{middle panel} shows
the evolution of the bar length. In the N-body case, the bar lengthens. This
occurs because as the pattern speed drops, bar-like orbits at larger radii are
possible. Stars are captured on these orbits, lengthening the bar. This
process does not occur in the SMUGGLE cases since the bar pattern speed is not
decreasing, and therefore the bar length remains constant. The \textit{lower
panel} shows the torque on the bar by different components. The solid lines
show the torque exerted by the halo in both the N-body and SMUGGLE cases. The
dashed line shows the torque exerted by the gas phase in the SMUGGLE run
(there is no gas in the N-body run). Details on how these properties
are calculated is given in the Methods section.}\label{fig:prop}
\end{figure}

\begin{figure*}[h]%
\centering
\includegraphics[width=0.9\textwidth]{fig/fig3.pdf}
\caption{The wake excited in the dark matter halo in the N-body case
(\textit{left panel}) and SMUGGLE case (\textit{right panel}) after
$2.6\,\textrm{Gyr}$. The \textit{left} and \textit{right panels} show a
surface density projection in the $x$-$y$ plane of the dark matter halo after
an axisymmetric average has been subracted. The solid line indicates the
direction of the bar while the dashed line indicates the direction of the halo
wake (both measured by taking the second Fourier component within a sphere of
all material within a radius of $X\,\textrm{kpc}$). The \textit{center panel}
shows the time evolution of the angle difference between the bar and the halo
wake, as measured from the second Fourier component. After the first Gyr, the
angle difference in the SMUGGLE case is smaller than in the N-body case by
about a factor of two, reflecting how the dark matter halo in the SMUGGLE case
is unable to exert as negative a torque on the bar as in the N-body
case.}\label{fig:wake}
\end{figure*}

\begin{figure}[h]
\centering
\includegraphics{fig/fig4.pdf}
\caption{The torque on the bar by the gas when the pattern speed of the bar is
kept at a constant value. Only gas within the corotation radius is able to
infall. Since slower bars have larger corotation radii, slower bars experience
a larger net torque than faster bars. The setup of the simulations used here
is identical to the SMUGGLE case discussed earlier and in the Methods section,
except the N-body disk is rotated as a solid body with a constant angular
velocity.}\label{fig:equil}
\end{figure}

@@

\item [Acknowledgements] Foo.

@@

  \item[Code Availability] Foo.

@@

\newpage

@@

\begin{center}
{\bf \Large \uppercase{Methods} }
\end{center}

@@

\noindent
{\bf Smooth Interstellar Medium Model}
\\
\noindent

In addition to the SMUGGLE model, we considered a simpler model of the
interstellar medium based upon ref.\cite{2003MNRAS.339..289S} In this model,
the multiphase nature of the interstellar medium is modeled in a subgrid
manner by allowing each resolution element to have a ``cold'' and ``hot''
component, with the equation of state of the gas suitably modified. Gas is
allowed to interchange between the ``cold'' and ``hot'' components through
processes such as cooling and stellar feedback. Cold gas is allowed to undergo
star formation. The pattern speed evolution of this setup is given in Extended
Data Fig.~X, which shows that the pattern speed evolves in a similar manner to
the SMUGGLE case shown in Fig.~\ref{fig:prop}.

@@

@@

\begin{figure}[h]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-vcirc.pdf}
\caption{The circular velocity curve for the N-body run (blue) and the SMUGGLE
run (orange) compared to observational estimates.\cite{2019ApJ...871..120E} We
see that the circular velocity curve for both runs is marginally larger than
the Milky Way's, but still comparable. The SMUGGLE circular velocity curve is
larger than the N-body curve due to the additional mass in the gas phase.}
\label{fig:vcirc}
\end{figure}

\begin{figure}[h]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-surf.pdf}
\caption{The time evolution of the gas surface density (\textit{upper}) and
the star formation rate (SFR) surface density (\textit{lower}) at various
times during the simulation. Colored lines indicate the profiles at selected
times during the simulation while the horizontal dashed lines indicate
approximate values for the Milky Way.\cite{2012ARAA..50..531K} We see that the
gas surface density is slightly higher but within an order of magnitude of the
Milky Way's typical gas surface density. We see a sharp decrease in the gas
and SFR surface density along the extent of the bar from $\sim1$ to
$\sim4\,\textrm{kpc}$, related to the gas inflow in this region. {\color{red}
I will add the actual points from ref.\cite{2012ARAA..50..531K} (Fig. 7), but
just placing approx horizontal lines for now. The profiles are roughly flat
anyways.}}
\label{fig:surf}
\end{figure}

@@

@@

\begin{figure}[h]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-A2.pdf}
\caption{The time evolution of the bar strength, measured as the second
Fourier component divided by the zeroth Fourier component (a formula is given
in the Methods section). We see that in the N-body case (blue) the bar
strength increases with time, consistent with previous results showing that
the bar strength increases as bars slow down. In the SMUGGLE case (orange), we
see that the bar strength remains roughly constant, possibly slightly
decreasing with time. This is also consistent with the expected relation
between pattern speed and strength since the bar in this case is not slowing
down.}
\label{fig:strength}
\end{figure}

@@

\begin{figure*}[h]%
\centering
\includegraphics[width=0.95\textwidth]{fig/fig-bar_decomp.pdf}
\caption{Demonstration of our bar decomposition procedure based on
ref.\cite{2016MNRAS.463.1952P} The \textit{left panel} shows a surface density
projection through the stellar component of the N-body simulation (disk and
bulge). The \textit{middle panel} shows the component of the disk identified
as being trapped in the bar while the \textit{right panel} shows the component
of the disk identified as not being trapped in the bar. The fact that the
untrapped stars form a roughly axisymmetric structure indicates our bar
decomposition is sufficiently accurate. {\color{red} I will remake this figure
- I ended up losing the files which stored the bar membership, so I just
pulled these screenshots from an earlier presentation.}}
\label{fig:decomp}
\end{figure*}

>>>>

\usepackage{epsfig,caption}
\usepackage{color}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{longtable}
% \usepackage{amssymb}
\usepackage{rotating,xcolor}
\usepackage{hyperref}
% \usepackage{tgbonum}
\usepackage[symbol]{footmisc}
\usepackage{placeins}

@@

\usepackage[misc]{ifsym}
\usepackage{xcolor}
\usepackage[labelfont=bf]{caption}

@@

\author{Angus Beane$^{1*}$, Lars Hernquist$^1$, Elena D'Onghia$^{2,3}$,
Federico Marinacci$^{4}$, Charlie Conroy$^{1}$, Jia Qi$^{5}$, Laura V.
Sales$^{6}$, Paul Torrey$^{5}$ and Mark Vogelsberger$^{7}$}

@@

\begin{affiliations}
% 1
\item Center for Astrophysics $|$ Harvard \& Smithsonian,  Cambridge, MA, USA.

% 2
\item Department of Physics, University of Wisconsin-Madison, Madison, WI, USA.

% 3
\item Department of Astronomy, University of Wisconsin-Madison, Madison, WI, USA.

% 4
\item Department of Physics \& Astronomy `Augusto Righi', University of Bologna, Bologna, Italy.

% 5
\item Department of Astronomy, University of Florida, Gainesville, FL, USA.

% 6
\item Department of Physics \& Astronomy, University of California, Riverside, CA, USA.

% 7
\item Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA.

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=0.8\textwidth]{fig/fig1.pdf}
\caption{Surface density projections in an N-body only simulation (upper
panels) and a simulation which includes the SMUGGLE model (bottom panels). We
can see that in the N-body run, the bar grows in length and strength. In the
SMUGGLE run (which includes a gaseous phase and a model for the multi-phase
interstellar medium), the bar remains at approximately the same length and
strength over the course of the simulation. The N-body model is identical to
the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

\begin{figure}[h!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig2.pdf}
\caption{Various bar properties from the N-body and SMUGGLE runs. The
\textit{upper panel} show the evolution of the pattern speed. As expected, the
bar in the N-body run slows down due to interactions between the bar and the
dark matter halo. However, the bar in the SMUGGLE run does not slow down and
instead remains at a constant pattern speed. The \textit{middle panel} shows
the evolution of the bar length. In the N-body case, the bar lengthens. This
occurs because as the pattern speed drops, bar-like orbits at larger radii are
possible. Stars are captured on these orbits, lengthening the bar. This
process does not occur in the SMUGGLE cases since the bar pattern speed is not
decreasing, and therefore the bar length remains constant. The \textit{lower
panel} shows the torque on the bar by different components. The solid lines
show the torque exerted by the halo in both the N-body and SMUGGLE cases. The
dashed line shows the torque exerted by the gas phase in the SMUGGLE run
(there is no gas in the N-body run). Details on how these properties
are calculated is given in the Methods section.}\label{fig:prop}
\end{figure}

\begin{figure*}[h!]%
\centering
\includegraphics[width=0.9\textwidth]{fig/fig3.pdf}
\caption{The wake excited in the dark matter halo in the N-body case
(\textit{left panel}) and SMUGGLE case (\textit{right panel}) after
$2.6\,\textrm{Gyr}$. The \textit{left} and \textit{right panels} show a
surface density projection in the $x$-$y$ plane of the dark matter halo after
an axisymmetric average has been subracted. The solid line indicates the
direction of the bar while the dashed line indicates the direction of the halo
wake (both measured by taking the second Fourier component within a sphere of
all material within a radius of $X\,\textrm{kpc}$). The \textit{center panel}
shows the time evolution of the angle difference between the bar and the halo
wake, as measured from the second Fourier component. After the first Gyr, the
angle difference in the SMUGGLE case is smaller than in the N-body case by
about a factor of two, reflecting how the dark matter halo in the SMUGGLE case
is unable to exert as negative a torque on the bar as in the N-body
case.}\label{fig:wake}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics{fig/fig4.pdf}
\caption{The torque on the bar by the gas when the pattern speed of the bar is
kept at a constant value. Only gas within the corotation radius is able to
infall. Since slower bars have larger corotation radii, slower bars experience
a larger net torque than faster bars. The setup of the simulations used here
is identical to the SMUGGLE case discussed earlier and in the Methods section,
except the N-body disk is rotated as a solid body with a constant angular
velocity.}\label{fig:equil}
\end{figure}

@@

The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics.\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R} Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar.\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C} Third
ed and a fast rotator, of which there is some
observational evidence.\cite{2019MNRAS.490.4740B} And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

@@

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which are exceptionally
gas-poor to be rotating very slowly. We therefore predict that observations
which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one such example of a galaxy which is known to be a slow rotator -- UGC
628.\cite{2009AA...499L..25C} This galaxy has been studied in detail by
ref.\cite{2016MNRAS.463.1751C}, who note that it indeed has a low gas
fraction. In general, we predict that all gas-rich spiral galaxies should
rotate quickly while gas-poor spiral galaxies should rotate slowly.

@@

\item [Acknowledgements] We would like to thank Greg L. Bryan, Drummond B.
Fielding, Keith Hawkins, Jason A. S. Hunt, Sarah M. R. Jeffreson, Kathryn V.
Johnston, Julio F. Navarro, David N. Spergel, Martin D. Weinberg, and Yanfei
Zou for helpful discussions. A.B. would like to thank Todd Phillips for
helpful discussions. Some computations in this paper were run on the FASRC
Cannon cluster supported by the FAS Division of Science Research Computing
Group at Harvard University. Resources supporting this work were also provided
by the NASA High-End Computing (HEC) Program through the NASA Advanced
Supercomputing (NAS) Division at Ames Research Center. A.B. was supported by
the Future Investigators in NASA Earth and Space Science and Technology
(FINESST) award number 80NSSC20K1536 during the completion of this work. L.H.
funding. E.D. funding. F.M. funding. C.C. funding. J.Q. funding. L.V.S.
funding. P.T. funding. M.V. funding.

@@

  \item[Code Availability] A public version of the AREPO code is available at
  \url{https://arepo-code.org/}.

@@

\FloatBarrier
\newpage

@@

\noindent
{\bf \Large \uppercase{Methods} }

\bigskip

@@

\noindent
{\bf Smooth Interstellar Medium Model}
\\
\noindent
In addition to the SMUGGLE model, we considered a simpler model of the
interstellar medium based upon ref.\cite{2003MNRAS.339..289S} In this model,
the multiphase nature of the interstellar medium is modeled in a subgrid
manner by allowing each resolution element to have a ``cold'' and ``hot''
component, with the equation of state of the gas suitably modified. Gas is
allowed to interchange between the ``cold'' and ``hot'' components through
processes such as cooling and stellar feedback. Cold gas is allowed to undergo
star formation. The pattern speed evolution of this setup is given in Extended
Data Fig.~X, which shows that the pattern speed evolves in a similar manner to
the SMUGGLE case shown in Fig.~\ref{fig:prop}.

@@

\begin{figure}[t]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-vcirc.pdf}
\caption{The circular velocity curve for the N-body run (blue) and the SMUGGLE
run (orange) compared to observational estimates.\cite{2019ApJ...871..120E} We
see that the circular velocity curve for both runs is marginally larger than
the Milky Way's, but still comparable. The SMUGGLE circular velocity curve is
larger than the N-body curve due to the additional mass in the gas phase.}
\label{fig:vcirc}
\end{figure}

@@

\begin{figure}[t!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-surf.pdf}
\caption{The time evolution of the gas surface density (\textit{upper}) and
the star formation rate (SFR) surface density (\textit{lower}) at various
times during the simulation. Colored lines indicate the profiles at selected
times during the simulation while the horizontal dashed lines indicate
approximate values for the Milky Way.\cite{2012ARAA..50..531K} We see that the
gas surface density is slightly higher but within an order of magnitude of the
Milky Way's typical gas surface density. We see a sharp decrease in the gas
and SFR surface density along the extent of the bar from $\sim1$ to
$\sim4\,\textrm{kpc}$, related to the gas inflow in this region. {\color{red}
I will add the actual points from ref.\cite{2012ARAA..50..531K} (Fig. 7), but
just placing approx horizontal lines for now. The profiles are roughly flat
anyways.}}
\label{fig:surf}
\end{figure}

@@

\begin{figure}[t!]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig-A2.pdf}
\caption{The time evolution of the bar strength, measured as the second
Fourier component divided by the zeroth Fourier component (a formula is given
in the Methods section). We see that in the N-body case (blue) the bar
strength increases with time, consistent with previous results showing that
the bar strength increases as bars slow down. In the SMUGGLE case (orange), we
see that the bar strength remains roughly constant, possibly slightly
decreasing with time. This is also consistent with the expected relation
between pattern speed and strength since the bar in this case is not slowing
down.}
\label{fig:strength}
\end{figure}

\begin{figure*}[t!]%
\centering
\includegraphics[width=0.95\textwidth]{fig/fig-bar_decomp.pdf}
\caption{Demonstration of our bar decomposition procedure based on
ref.\cite{2016MNRAS.463.1952P} The \textit{left panel} shows a surface density
projection through the stellar component of the N-body simulation (disk and
bulge). The \textit{middle panel} shows the component of the disk identified
as being trapped in the bar while the \textit{right panel} shows the component
of the disk identified as not being trapped in the bar. The fact that the
untrapped stars form a roughly axisymmetric structure indicates our bar
decomposition is sufficiently accurate. {\color{red} I will remake this figure
- I ended up losing the files which stored the bar membership, so I just
pulled these screenshots from an earlier presentation.}}
\label{fig:decomp}
\end{figure*}

@@

@@
